\sys: Light-touch Data-Parallel Shell Processing

### 2020-05-22 pldi/draft.tex
Pass over DFG

The core of \sys is a dataflow graph model that is expressive enough
to support a large subset of shell programs. A fundamental difference
with other dataflow graph models is that in \sys's model contains
information about the order in which a node reads its inputs. This
enables a set of graph transformations that expose data parallelism as
task parallelism. These transformations can then be applied in an
iterative way, optimizing the program to utilize available
computational resources.

%% The core of \sys is a dataflow graph model, inspired by the ones used
%% in popular distributed stream processing systems~\sx{related} \kk{Is
%%   this reference here necessary?}. The dataflow model can express a
%% large subset of Shell programs and inherently supports the data
%% parallelism found in shell pipelines by exposing it as task
%% parallelism. More preciely, we have developed a set of
%% parallelization-exposing optimizations represented as
%% semantics-preserving graph transformations. These transformations can
%% then be applied in an iterative way, optimizing the program to utilize
%% available computational resources.

@@

The two main shell abstractions are (i) files (or pipes, \emph{viz.}
correspondence between files and pipes in \S\ref{bg:pipelines}) that
contain data, and (ii) commands that communicate through these
files. Based on that, in \sys's graph model nodes represent commands
and edges represent files.

We first introduce basic notation that will be useful later on.  For a
set $D$, we write $D\kstar$ to denote the set of all finite words over
$D$. For words $x, y \in D\kstar$, we write $x \cdot y$ or $xy$ to
denote their concatenation. We write $\eps$ for the empty word. We say
that $x$ is a \emph{prefix} of $y$, and we write $x \leq y$, if there
is a word $z$ such that $y = xz$. The $\leq$ order is reflexive,
antisymmetric, and transitive (i.e. is a partial order) and is often
called the \emph{prefix order}.

\heading{Edges -- Files}

Edges in the dataflow graph represent files, the basic data
abstraction of the shell. They are used as communication channels
between nodes in the graph, and as the input or output of the whole
graph. Edges are represented as possibly unbounded streams of type
$D\kstar$. Edges can either refer to named files or FIFO pipes used
for interprocess communication. Edges that do not start from a node in
the graph represent the graph input; edges that do not point to a node
in the graph are its outputs.

@@

\heading{Nodes -- Commands}

A node $f$ of the graph represents a function from a (possibly empty)
list of inputs
%% streams
to a list of outputs
%% streams
$f : [D\kstar] \to [D\kstar]$, where $D$ represents the basic data
type of a line of characters.
%% \km{Is the number of input/output
%%   streams variable? If $m$ is the number of input streams, is the
%%   number of output streams a function of $m$? If this the case, then
%%   it is not captured by this type. Maybe it would help to show a
%%   couple of example operators together with the types of their
%%   denotations. A graphical representation (as in
%%   Fig.~\ref{fig:parallelization-transformation} would also be nice.}
%% \kk{Is there a reason to restrict the number of outputs to be a
%%   function of the number of inputs? I think that we don't need such a
%%   restriction, therefore the above type should be fine.}
This representation captures \todo{the majority/all} of commands in
the \tsta, \tpur, \tnpu. Note that these functions should only produce
output in the form of files and not perform any other side effect,
such as sending signals.  We require that the function $f$ is monotone
w.r.t.\ a lifting of the prefix order for a sequence of inputs
%% \kk{Formalize
%%   this lifting if essential for later development}.
%% More formally, $\forall f, x, y, f(x) = y$, then for any $x'$, there
%% exists $y'$ such that $f(x.x') = y.y'$, where $.$ is standard
%% concatenation and will be omitted when obvious.
This captures the idea that a node cannot retract output that it has
already produced.

\heading{\todo{Streaming} Commands}

\todo{Streaming} commands consume their inputs sequentially and one
element at a time, with the possible exception of static input files,
and produce one output file. The quintessential example of
\todo{streaming} commands is \ttt{cat}, that consumes its inputs in
order, producing their concatenation as output. A more interesting
example is \ttt{comm -23 f1 f2}, which has \ttt{f2} as its static
input (it needs to read it completely before consuming anything from
\ttt{f1} and then consumes \ttt{f1} one element at a time. These
commands can be represented as functions of type $f : D\kstar \times
[D\kstar] \to D\kstar$, where the first argument represents the
concatenation of the inputs that are consumed one at a time and the
second argument represents the static inputs.
%% \km{Again, a graphical
%%   presentation of these operators would be useful. Maybe with a way to
%%   distinguish between between streaming inputs and static inputs? Btw,
%%   this distinction is not completely clear to me at this
%%   point. Clearly, not all programs that we can write are streaming
%%   (monotone).} \kk{Do you mean that there are shell scripts that are
%%   non monotone? That is true but we are assuming monotone nodes. Do I
%%   miss something?}
We focus on these commands because a large subset of \sta and \pur
commands fall in this category.

@@

%% \kk{If we have time and space we could give an example of a script
%%   together with its dataflow graph to give more intuition about nodes
%%   and streaming commands.}

@@

%% \kk{Move this assumption in the transformation: Nodes may read from
%%   more files than just the files of their input stream (resp. write to
%%   more files than their output stream).  However, these files are
%%   static (\eg used for configuration or logging) and do not represent
%%   streams of data, and therefore are not considered part of the
%%   dataflow graph.  For now, we assume that these static files can only
%%   be accessed by one node, thus ensuring that they do not interfere
%%   with the distributed implementation.
%% %% Because of the assumption in section Front End \ref{}, that each file
%% %% exists only once in the dataflow graph, we can safely assume that
%% %% reading and writing to these static files does not interfere with the
%% %% distributed implementation, and doesn't alter the behaviour of the
%% %% program.
%% }

We have defined a set of semantics-preserving graph transformations
that act as parallelization-exposing optimizations. Both the domain
and range of these transformations is a graph in our dataflow model
and therefore they can be composed arbitrarily and in any order.
%% Having a dataflow graph representation allows us to define
%% parallelization optimizations as graph transformations.
Before describing the different types of optimizations, we formalize
the intuition about stateless and pure commands that was described
earlier~\sx{cmd}.

@@

\heading{Stateless and Parallelizable Pure Commands}

As mentioned in \Cref{cmd}, stateless commands such as \ttt{tr}
operate independently on individual elements of the stream
(characters, lines, or files) without maintaining any state. In this
work, we consider lines as the data quantum, so only commands that are
stateless regarding lines (or any finer granularity such as
characters) are considered stateless. Formally, a
\todo{streaming} command $f$ is stateless if it commutes with
the operation of concatenation, \ie it is a semigroup homomorphism:
%% with respect to concatenation \km{It commutes with the operation of concatenation. This is the definition of (semigroup) homomorphism. I am not sure that linearity is an appropriate term.}
%% , \ie satisfying the following equation:

@@

Similarly, some pure commands (such as \ttt{sort} and \ttt{wc}) can be
parallelized using divide-and-conquer techniques. More formally, these
pure commands $f$ can be implemented as a combination of functions map
$m$ and an associative fold $f'$, satisfying the following equation:

@@

\noindent
This means that we get the same output by applying $f$ to a
concatenation of two inputs $x, x'$, or by applying the fold function
$f'$ to the concatenation of the outputs produced by applying map $m$
to each of these inputs.
%% \km{We had also discussed here the case of
%%   operators that can be described as associative and commutative
%%   functions on the data (e.g., sum). In this case the parallelization
%%   has the flavor of data parallelism: one can split the input stream
%%   in an arbitrary way (e.g., randomly or using some other
%%   load-balancing scheme) and merge the outputs. For the more broad
%%   class of associative operations (e.g., sort), the formulation of
%%   this paragraph makes perfect sense. I just want to mention that this
%%   feels a lot more like \textbf{batching} than map-reduce-like data
%%   parallelization.} \kk{Thanks, I think you are right. Batching is a
%%   better name for it. In the end we didn't handle
%%   commutative-associative function (that are more like map-reduce) and
%%   so we changes the name to map-fold to not imply commutativity.}

\heading{Parallelization Transformations}
%
Based on these equations, we can define a \emph{node parallelization
  transformation} on a stateless node $v$ that is preceded by a
\ttt{cat} with $n$ input streams and is followed by a node $v'$. The
transformation replaces $v$ with $n$ new nodes, routing each of the
$n$ input streams to one of them, and commutes the \ttt{cat} node
after them to concatenate their outputs and transfer them to
$v'$. Since each incoming edge represents a stream of data $x_i :
D^*$, and the only behaviour of a dataflow graph is its output, this
optimization $v(x_1 \cdot x_2 \cdots x_n, s) \Rightarrow v(x_1, s)
\cdot v(x_2, s) \cdots v(x_n, s)$ can be shown to preserve the
behavior of the graph. Furthermore, it can be straightforwardly
extended to pure commands that can be implemented by a map-fold pair
$(m, r)$ as $ v(x_1 \cdot x_2 \cdots x_n, s) \Rightarrow r(m(x_1, s)
\cdot m(x_2, s) \cdots m(x_n, s), s)$. \todo{A visual representation
  of this transformation can be seen in}
\Cref{fig:parallelization-transformation}.

@@

\heading{Auxiliary Transformations}

In addition to the above, we also define a set of auxiliary
transformations that are shown in
\Cref{fig:auxiliary-transformations}.  The first two enable the
parallelization transformations by inserting \ttt{cat} nodes, either
by using them to concatenate many inputs of one command, or by first
adding a \ttt{split} node that splits the input in batches. The final
one inserts a relay that performs the identity transformation on its
input. Relay nodes can be useful for monitoring and debugging, as well
as for performance improvements, as we will see in \Cref{optimizer}.

>>>>

The core of \sys is a dataflow graph (DFG) model that is expressive enough to support a large subset of shell programs.\nv{Immediate Q: How Large? Which programs are not supported? Why?}
A fundamental difference with other DFG models is that it contains information about the order in which a node reads its inputs.
This enables a set of graph transformations that expose data parallelism as task parallelism.\nv{What does this mean? What if they are fundamentally different in the reivewer's mind? Example?}
These transformations can then be applied in an iterative way, optimizing the program to utilize available computational resources.

%% The core of \sys is a DFG model, inspired by the ones used
%% in popular distributed stream processing systems~\sx{related} \kk{Is
%%   this reference here necessary?}. The dataflow model can express a
%% large subset of Shell programs and inherently supports the data
%% parallelism found in shell pipelines by exposing it as task
%% parallelism. More preciely, we have developed a set of
%% parallelization-exposing optimizations represented as
%% semantics-preserving graph transformations. These transformations can
%% then be applied in an iterative way, optimizing the program to utilize
%% available computational resources.

@@

The two main shell abstractions are
  (i) files (or pipes, \emph{viz.} correspondence between files and pipes in \S\ref{bg:pipelines}\nv{Not in this version. Should we call them "streams"?}) that contain data, and
  (ii) commands that communicate through these files.
\nv{quick solution: change all files to streams and make the equivalence explicit here}
\sys's DFG model represents commands as nodes and \todo{streams} as edges.

We first introduce basic notation used throughput this section.
For a set $D$, we write $D\kstar$ to denote the set of all finite words over $D$.
For words $x, y \in D\kstar$, we write $x \cdot y$ or $xy$ to denote their concatenation.
We write $\eps$ for the empty word.
We say that $x$ is a \emph{prefix} of $y$, and we write $x \leq y$, if there is a word $z$ such that $y = xz$.
The $\leq$ order is reflexive, antisymmetric, and transitive (\ie is a partial order) and is often called the \emph{prefix order}.

\heading{Edges---Streams}
Edges in the DFG represent streams, the basic data abstraction of the shell.
They are used as communication channels between nodes in the graph, and as the input or output of the entire graph.
Edges are represented as possibly unbounded streams of type $D\kstar$.
Edges can either refer to named files or FIFO pipes used for interprocess communication.
Edges that do not start from a node in the graph represent the graph input;
  edges that do not point to a node in the graph are its outputs.
\nv{OK This distinction is explained in this par; I would put this earlier}

@@

\heading{Nodes---Commands}
A node $f$ of the graph represents a function from a (possibly empty)
list of inputs
%% streams
to a list of outputs
%% streams
$f : [D\kstar] \to [D\kstar]$, where $D$ represents the basic data
type of a line of characters.
%% \km{Is the number of input/output
%%   streams variable? If $m$ is the number of input streams, is the
%%   number of output streams a function of $m$? If this the case, then
%%   it is not captured by this type. Maybe it would help to show a
%%   couple of example operators together with the types of their
%%   denotations. A graphical representation (as in
%%   Fig.~\ref{fig:parallelization-transformation} would also be nice.}
%% \kk{Is there a reason to restrict the number of outputs to be a
%%   function of the number of inputs? I think that we don't need such a
%%   restriction, therefore the above type should be fine.}
This representation captures all the commands in
the \tsta, \tpur, and \tnpu. Note that these functions should only produce
output in the form of files and not perform any other side effect,
such as sending signals.  We require that the function $f$ is monotone
with respect to a lifting of the prefix order for a sequence of inputs
%% \kk{Formalize
%%   this lifting if essential for later development}.
%% More formally, $\forall f, x, y, f(x) = y$, then for any $x'$, there
%% exists $y'$ such that $f(x.x') = y.y'$, where $.$ is standard
%% concatenation and will be omitted when obvious.
This captures the idea that a node cannot retract output that it has
already produced.

\heading{\todo{Streaming} Commands}
These commands consume their inputs sequentially and one
element at a time, with the possible exception of static input files,
and produce one output file. The quintessential example of
\todo{streaming} commands is \ttt{cat}, that consumes its inputs in
order, producing their concatenation as output. A more interesting
example is \ttt{comm -23 f1 f2}, which has \ttt{f2} as its static
input (it needs to read it completely before consuming anything from
\ttt{f1} and then consumes \ttt{f1} one element at a time. These
commands can be represented as functions of type $f : D\kstar \times
[D\kstar] \to D\kstar$, where the first argument represents the
concatenation of the inputs that are consumed one at a time and the
second argument represents the static inputs.
%% \km{Again, a graphical
%%   presentation of these operators would be useful. Maybe with a way to
%%   distinguish between between streaming inputs and static inputs? Btw,
%%   this distinction is not completely clear to me at this
%%   point. Clearly, not all programs that we can write are streaming
%%   (monotone).} \kk{Do you mean that there are shell scripts that are
%%   non monotone? That is true but we are assuming monotone nodes. Do I
%%   miss something?}
We focus on these commands because a large subset of \sta and \pur
commands fall in this category.
\nv{Based on the intro of 4 or 4.1 I did not expect streaming. I understand edges and nodes, but how do streaming commands fit into the picture? the answer might be the last sentence which should be first---reviewer: why am I reading what I'm reading?:P}

@@

%% \kk{If we have time and space we could give an example of a script
%%   together with its DFG to give more intuition about nodes
%%   and streaming commands.}

@@

%% \kk{Move this assumption in the transformation: Nodes may read from
%%   more files than just the files of their input stream (resp. write to
%%   more files than their output stream).  However, these files are
%%   static (\eg used for configuration or logging) and do not represent
%%   streams of data, and therefore are not considered part of the
%%   DFG.  For now, we assume that these static files can only
%%   be accessed by one node, thus ensuring that they do not interfere
%%   with the distributed implementation.
%% %% Because of the assumption in section Front End \ref{}, that each file
%% %% exists only once in the DFG, we can safely assume that
%% %% reading and writing to these static files does not interfere with the
%% %% distributed implementation, and doesn't alter the behaviour of the
%% %% program.
%% }

\sys defines a set of semantics-preserving graph transformations
that act as parallelization-exposing optimizations. Both the domain
and range of these transformations is a graph in our DFG model
and therefore they can be composed arbitrarily and in any order.
%% Having a DFG representation allows us to define
%% parallelization optimizations as graph transformations.
Before describing the different types of optimizations, we formalize
the intuition behind classes \sta and \pur described informally
earlier~\sx{cmd}.

@@

\heading{Stateless and Parallelizable Pure Commands}
As mentioned in \Cref{cmd}, stateless commands such as \ttt{tr}
operate independently on individual elements of the stream
(characters, lines, or files) without maintaining any state. In this
work, we consider lines as the data quantum, so only commands that are
stateless regarding lines (or any finer granularity such as
characters) are considered stateless. Formally, a
\todo{streaming} command $f$ is stateless if it commutes with
the operation of concatenation, \ie it is a semigroup homomorphism:
%% with respect to concatenation \km{It commutes with the operation of concatenation. This is the definition of (semigroup) homomorphism. I am not sure that linearity is an appropriate term.}
%% , \ie satisfying the following equation:

@@

Similarly, some pure commands (such as \ttt{sort} and \ttt{wc}) can be
parallelized using divide-and-conquer techniques. More formally, these
pure commands $f$ can be implemented as a combination of a function map
$m$ and an associative fold $f'$, satisfying the following equation:

@@

\noindent
This means that we get the same output by applying $f$ to a
concatenation of two inputs $x, x'$ as by applying the fold function
$f'$ to the concatenation of the outputs produced by applying map $m$
to each of $x$ and $x'$.
%% \km{We had also discussed here the case of
%%   operators that can be described as associative and commutative
%%   functions on the data (e.g., sum). In this case the parallelization
%%   has the flavor of data parallelism: one can split the input stream
%%   in an arbitrary way (e.g., randomly or using some other
%%   load-balancing scheme) and merge the outputs. For the more broad
%%   class of associative operations (e.g., sort), the formulation of
%%   this paragraph makes perfect sense. I just want to mention that this
%%   feels a lot more like \textbf{batching} than map-reduce-like data
%%   parallelization.} \kk{Thanks, I think you are right. Batching is a
%%   better name for it. In the end we didn't handle
%%   commutative-associative function (that are more like map-reduce) and
%%   so we changes the name to map-fold to not imply commutativity.}

\heading{Parallelization Transformations}
Based on these equations, we can define a \emph{node parallelization
  transformation} on a stateless node $v$ that is preceded by a
\ttt{cat} with $n$ input streams and is followed by a node $v'$ (Fig.~\ref{fig:parallelization-transformation}). The
transformation replaces $v$ with $n$ new nodes, routing each of the
$n$ input streams to one of them, and commutes the \ttt{cat} node
after them to concatenate their outputs and transfer them to
$v'$. Since each incoming edge represents a stream of data $x_i :
D^*$, and the only behaviour of a DFG is its output, this
optimization $v(x_1 \cdot x_2 \cdots x_n, s) \Rightarrow v(x_1, s)
\cdot v(x_2, s) \cdots v(x_n, s)$ can be shown to preserve the
behavior of the graph.

This transformation can extended straightforwardly to pure commands implemented by a map-fold pair
$(m, r)$ as $ v(x_1 \cdot x_2 \cdots x_n, s) \Rightarrow r(m(x_1, s)
\cdot m(x_2, s) \cdots m(x_n, s), s)$.
% \todo{It is shown in Fig.}\ref{fig:parallelization-transformation}.

@@

\heading{Auxiliary Transformations}
\todo{We} also define a set of auxiliary transformations that are shown in
\Cref{fig:auxiliary-transformations}.  The first two \nv{Name them?} enable the
parallelization transformations by inserting \ttt{cat} nodes,\footnote{\nv{This is minor, but for the model maybe name them concatenation nodes or ++? rather than \ttt{cat}}} either
by using them to concatenate many inputs of one command, or by first
adding a \ttt{split} node that splits the input in batches. The final
one inserts a relay that performs the identity transformation on its
input. Relay nodes can be useful for monitoring and debugging, as well
as for performance improvements~\sx{optimizer}.
